\documentclass[english,bibfile=bibliography]{algotemplate/thesis}

\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}

% ------------------------------------------ Title page ------------------------
\title{Time-Dependent Decision Making in Heuristic Chess Search}
\type{Bachelor Thesis}
\author{Marvin Becker}
\studentno{01/1365135}
\group{Algorithmics}
\department{Computer and Information Science}
\institute{University of Konstanz}
\advisor{Prof.~Dr.~Sabine Storandt}
\reviewer{Prof.~Dr.~Sabine Storandt\\ & Prof.~Dr.~Michael Grossniklaus}
\date{2026-09-01}
% ------------------------------------------------------------------------------

% ------------------------------------------ Abstract --------------------------
\begin{abstract}
Chess engines rely heavily on heuristic search techniques to cope with the
exponential growth of the game tree. Among these techniques, iterative deepening
combined with alpha-beta pruning has become a standard approach due to its
ability to produce strong moves under strict time constraints.

This thesis presents the design and implementation of a modular chess engine and
investigates the relationship between search time and decision quality. The
central question is where the sweet spot between invested time and move quality
lies: how much additional search time actually translates into better decisions,
and at which point further computation no longer changes the chosen move.

The engine is implemented from scratch in \texttt{C++} using a classical
mailbox-based board representation. Its baseline integrates established
techniques such as alpha-beta pruning, transposition tables, quiescence search,
heuristic move ordering, and a two-stage time management system with soft and
hard deadlines. On top of this baseline, the thesis develops two further search
improvements: Principal Variation Search (PVS), which verifies all but the
presumed best move with inexpensive null-window searches, and the reuse of
search trees from previous searches, realized through a persistent
transposition table that allows the engine to resume earlier computations.

The experimental evaluation studies time-dependent decision making along four
axes: move stability, i.e., the depth at which the best move stabilizes and no
longer changes with deeper search; time-to-quality curves, which capture how
move quality evolves under increasing time limits; depth-versus-time behavior
with and without Principal Variation Search; and a PVS parameter sweep that
determines which combination of scouting depth and number of fully searched
moves yields the best ratio of cost reduction to move quality.

The findings aim to characterize how gracefully decision quality degrades under
tight time budgets and to identify configurations in which a deliberately simple
engine architecture converts limited search time into stable, high-quality
decisions.
\end{abstract}
% ---------------
% ------------------------------------------------------------------------------

% --------------------------------- Document / Main Content --------------------
\begin{document}

\input{body}

\end{document}