\section{Related Work}
\label{sec:related_work}
% Classical chess engines, alpha-beta, iterative deepening, time control

The problem of efficient search in chess engines has been studied extensively
over several decades. Early approaches were based on the Minimax algorithm,
which provides an optimal decision strategy under the assumption of perfect
play~\cite{russell2010aima}. However, due to the exponential growth of the game tree, Minimax quickly becomes infeasible even at moderate depths.

Alpha-Beta pruning was introduced as a major improvement over plain Minimax,
allowing large parts of the search tree to be pruned without affecting the
correctness of the result~\cite{knuth1975alphabeta}. Subsequent work showed that
the effectiveness of Alpha-Beta pruning depends crucially on the order in which
moves are explored. With perfect move ordering, the effective branching factor can be reduced dramatically, enabling much deeper searches within the same computational budget~\cite{marsland1986review}.

Iterative deepening has emerged as a standard technique to improve move ordering
and to support time-controlled search. By repeatedly searching the game tree with
increasing depth limits, iterative deepening ensures that a valid best move is
always available while simultaneously refining the move ordering for deeper
iterations. Despite the apparent redundancy caused by re-searching parts of the
tree, it has been shown that the overhead is often compensated by improved
pruning efficiency~\cite{marsland1986review}.

A detailed analysis of Alpha-Beta pruning efficiency in chess engines is
presented by Nair~\cite{nair_improving_ab}. The paper investigates how different
move ordering heuristics and search enhancements affect the pruning behavior of
Alpha-Beta search. In particular, it highlights the strong interaction between
iterative deepening, transposition tables, and heuristic move ordering. The
results indicate that improved move ordering can significantly reduce the number
of explored nodes, even when additional search iterations are introduced.

Transposition tables have further improved search efficiency by allowing engines
to reuse previously computed results for positions reached via different move
orders~\cite{zobrist1970hashing}. By storing evaluations together with depth
information and bound types, transposition tables enable both direct cutoffs and
enhanced move ordering. Their effectiveness is especially pronounced in
combination with iterative deepening, where information from earlier iterations
can be reused almost immediately~\cite{warnock1988search}.

Quiescence search addresses another well-known limitation of depth-limited
search, commonly referred to as the horizon effect. By extending the search in
tactically unstable positions, quiescence search produces more reliable
evaluations and has become a standard component of classical chess
engines~\cite{marsland1986review}.

While modern state-of-the-art engines incorporate additional techniques such as
bitboards, evaluation learning, and neural networks~\cite{silver2018alphazero},
the classical search principles discussed above remain fundamental. This work
builds upon these established ideas and focuses specifically on the interaction
between iterative deepening and time constraints, using a deliberately simple
and transparent engine architecture to isolate and analyze their effects.