\section{Experimental Evaluation}
\label{sec:evaluation}

This section presents the experimental evaluation of the implemented chess
engine. The evaluation pursues the central question of this thesis: where lies
the sweet spot between invested search time and decision quality? To this end,
it first establishes a reliable baseline for the playing strength of the engine
and quantifies the effect of the search optimizations, and then studies
time-dependent decision making directly through move stability, time-to-quality,
and depth-versus-time measurements, including the influence of Principal
Variation Search and its tuning parameters.

\subsection{Research Objectives}
\label{subsec:research_objectives}

The experimental evaluation addresses the following research questions:

\begin{itemize}
    \item How strong is the optimized engine under realistic tournament time
    controls when compared to a fixed reference opponent?
    \item How large is the performance difference between optimized search and
    plain Negamax with Alpha-Beta pruning when evaluated on identical positions?
    \item \textbf{Move stability:} at which search depth does the best move
    stabilize, so that deeper search no longer changes the decision?
    \item \textbf{Time-to-quality:} how does move quality change as the time
    limit is increased, and where does additional time stop paying off?
    \item \textbf{Depth versus time:} how does the reachable depth per time
    budget differ with and without Principal Variation Search?
    \item \textbf{PVS parameter sweep:} which combination of minimum scouting
    depth (\texttt{PvsMinDepth}) and number of fully searched moves
    (\texttt{PvsScoutAfterMove}) yields the best ratio of cost reduction to
    move quality?
\end{itemize}

The first two questions establish the baseline strength and the effect of the
search optimizations; the remaining four questions form the core of this thesis
and characterize time-dependent decision making directly.

\subsection{Engine-versus-Engine Evaluation}
\label{subsec:engine_eval}

\subsubsection{Experimental Setup}
\label{subsubsec:engine_setup}

The engine under evaluation (\emph{Helix}) was tested in head-to-head matches
against a fixed reference opponent under controlled and reproducible conditions.

\paragraph{Engines.}
Helix was evaluated against a development version of Stockfish
(\texttt{Stockfish dev-20251117-035cb146}). To obtain a meaningful comparison in
the target strength range, Stockfish was configured using its built-in Elo
limitation mechanism with \texttt{UCI\_LimitStrength} enabled and a target
strength of 2000 Elo. In addition, Stockfish was restricted to a single search
thread and a hash size of 32\,MB. Pondering was disabled for both engines.

\paragraph{Time Control.}
All games were played using a classical tournament time control of 5 minutes per
side with an increment of 0.5 seconds per move. This setting ensures that time
management and iterative deepening behavior have a significant influence on
search quality while still allowing sufficiently deep searches.

\paragraph{Match Configuration.}
A total of 400 games were played using the CuteChess testing framework. No opening
book was used; all games started from the standard initial position. Colors were
alternated, and each engine played an equal number of games as White and Black to
ensure fairness.

\paragraph{Hardware and Software Environment.}
All experiments were conducted on a MacBook Pro equipped with an Apple M3 Pro
processor and 18\,GB of RAM. Both engines were compiled in release mode and run as
single-threaded processes. Apart from the explicitly stated options, all engine
parameters were kept constant throughout the experiments.

\paragraph{Evaluation Methodology.}
Playing strength was measured using Elo difference estimates derived from match
results. In addition to the Elo estimate, the likelihood of superiority (LOS) and
the draw ratio were recorded to assess statistical confidence and game
decisiveness.

\subsubsection{Results}
\label{subsubsec:engine_results}

Table~\ref{tab:elo_results} summarizes the results of the match series between
Helix and the reference Stockfish configuration.

\begin{table}[ht]
\centering
\begin{tabular}{@{}lr@{}}
\toprule
Metric & Value \\ \midrule
Games played & 400 \\
Elo difference & $+161.9 \pm 36.2$ \\
Likelihood of Superiority (LOS) & $100.0\,\%$ \\
Draw ratio & $7.5\,\%$ \\ \bottomrule
\end{tabular}
\caption{Match results of Helix against Stockfish limited to 2000 Elo.}
\label{tab:elo_results}
\end{table}

Out of the 400 games played, Helix achieved a clear majority of wins. The low draw
ratio indicates that most games were decided decisively rather than through
repetition or stalemate.

A breakdown of decisive outcomes shows that Helix won 279 games by checkmate
(130 as Black and 149 as White), while Stockfish won 91 games by checkmate
(38 as Black and 53 as White). The remaining 30 games ended in draws due to
threefold repetition.

The estimated Elo difference of approximately $+162$ points in favor of Helix is
statistically significant, as indicated by the $100\,\%$ likelihood of
superiority. It should be emphasized that the reference engine was deliberately
limited to 2000 Elo; the reported difference therefore reflects relative
performance under this configuration rather than an absolute strength comparison.

\subsection{Search Optimization Ablation Study}
\label{subsec:ablation}

To isolate the effect of the implemented search optimizations, an additional
controlled ablation experiment was conducted. In contrast to the
engine-versus-engine evaluation, this experiment focuses exclusively on search
efficiency under identical time constraints.

\subsubsection{Experimental Design}

The engine was evaluated in two configurations:
\begin{itemize}
    \item \textbf{Optimized search:} Full implementation including transposition
    tables and heuristic move ordering (TT move, captures, killer moves, history
    heuristic).
    \item \textbf{Unoptimized search:} Plain Negamax with Alpha-Beta pruning,
    without transposition tables and without heuristic move ordering. Moves are
    explored in their generated order.
\end{itemize}

Both configurations use the same board representation, move generator, and
evaluation function. Any observed differences can therefore be attributed
directly to the presence or absence of search optimizations.

\subsubsection{Positions and Time Control}

The experiment was performed on a fixed set of 128 predefined chess positions.
For each position, both engine configurations were given a fixed search time
limit of 2 seconds. No depth or node limits were imposed.

This setup ensures that both configurations operate under identical external
constraints and that observed differences reflect how efficiently the available
time is converted into search depth and explored nodes.

\subsubsection{Evaluation Metrics}

For each position and configuration, the following metrics were recorded:
\begin{itemize}
    \item total number of explored nodes,
    \item maximum search depth reached within the time limit,
    \item average search depth over all positions.
\end{itemize}

All results are aggregated across the full set of 128 positions.

\subsubsection{Results}

Table~\ref{tab:ablation_nodes} compares the average number of explored nodes per
search depth for the optimized and unoptimized configurations under the fixed
2-second time limit.

\begin{table}[ht]
\centering
\begin{tabular}{@{}rcc@{}}
\toprule
Depth & Optimized Nodes (avg.) & Unoptimized Nodes (avg.) \\ \midrule
1 & $\approx 36$ & $\approx 1651$ \\
2 & $\approx 114$ & $\approx 278$ \\
3 & $\approx 625$ & $\approx 2596$ \\
4 & $\approx 1689$ & $\approx 19766$ \\
5 & $\approx 9410$ & $\approx 185529$ \\
\bottomrule
\end{tabular}
\caption{Average number of explored nodes per depth under a fixed 2-second time
limit, aggregated over 128 positions.}
\label{tab:ablation_nodes}
\end{table}

The optimized configuration consistently reaches greater search depths and
explores significantly fewer nodes per depth level. This indicates a substantial
reduction in the effective branching factor compared to the unoptimized search.

In a subset of the tested positions, both configurations selected identical best
moves and principal variations despite differences in achieved search depth. A
detailed discussion of this observation is provided in
Section~\ref{sec:discussion}.