\section{Introduction}
\label{sec:introduction}
% Motivation, problem statement, goals, research question

Chess has long been regarded as a canonical benchmark problem in artificial
intelligence and algorithm engineering. Despite its simple and fully specified
rules, the game exhibits an enormous search space caused by its high branching
factor and strategic depth. As a consequence, exhaustive search of the complete
game tree is computationally infeasible, even with modern hardware, and chess
engines must rely on heuristic search techniques to make strong decisions under
limited computational resources.

At the core of most classical chess engines lies a depth-limited game tree
search based on Minimax or Negamax, enhanced by Alpha-Beta pruning. Over the
years, a variety of improvements have been proposed to increase search
efficiency, including iterative deepening, transposition tables, quiescence
search, and heuristic move ordering. Among these techniques, iterative
deepening has become a standard approach, particularly in time-controlled
settings, as it allows the engine to gradually refine its decision while always
maintaining a valid best move.

The primary motivation of this work is to understand how a chess engine should
spend its limited time. More search time generally yields deeper searches and
better moves, but the relationship is far from linear: at some point the chosen
move stabilizes and additional computation no longer changes the decision. At
the same time, search refinements such as Principal Variation Search and the
reuse of previous search trees change how efficiently a given time budget is
converted into depth and decision quality. This leads to the central research
question of this thesis:

\emph{Where lies the sweet spot between invested search time and decision
quality in heuristic chess search, and how do search refinements such as
Principal Variation Search and search tree reuse shift this trade-off?}

To make this question measurable, the thesis examines four concrete aspects:
(i) \emph{move stability}, i.e., the depth at which the best move stabilizes
and deeper search no longer changes the decision, (ii) the
\emph{time-to-quality curve}, i.e., how move quality changes as the time limit
increases, (iii) \emph{depth-versus-time} behavior with and without Principal
Variation Search, and (iv) a \emph{PVS parameter sweep} that determines which
combination of minimum scouting depth and number of fully searched moves yields
the best ratio of cost reduction to move quality.

To address this question, a complete chess engine is implemented from scratch in
\texttt{C++}. The engine employs a classical mailbox-based board representation
rather than highly optimized bitboard techniques, which keeps the impact of
search algorithms and time management observable in isolation from low-level
optimizations; the underlying design rationale is discussed in
\cref{sec:architecture}.

The implemented engine integrates a Negamax search with Alpha-Beta pruning,
iterative deepening, quiescence search, transposition tables, and heuristic move
ordering. A dedicated time management module supports multiple search modes,
including fixed depth, fixed time, node-limited, and infinite search, following
the UCI protocol. In particular, a two-stage time budget with soft and hard
deadlines is used to control the interaction between iterative deepening and
time constraints.

Beyond this baseline, this thesis develops two further search improvements:
Principal Variation Search and the reuse of search trees from previous searches
through a persistent transposition table. Both techniques are introduced
formally in \cref{sec:preliminaries} and described at the implementation level
in \cref{sec:algorithm_implementation}.

The contribution of this work is threefold. First, it provides a clean and
modular implementation of a classical chess engine that serves as a transparent
experimental platform. Second, it extends the baseline search with Principal
Variation Search and transposition-based search tree reuse and describes how
these techniques are realized in practice. Third, it presents an empirical
evaluation of time-dependent decision making, analyzing move stability across
depths, time-to-quality curves, depth-versus-time behavior with and without
PVS, and the effect of the PVS tuning parameters. The results aim to locate the
practical sweet spot between search time and decision quality in time-controlled
chess engines.

The remainder of this thesis is structured as follows. \cref{sec:related_work}
discusses relevant prior work in computer chess and game tree search.
\cref{sec:preliminaries} introduces the necessary theoretical background.
\cref{sec:algorithm_implementation} describes the architecture of the engine and
the implementation of the baseline search, the Principal Variation Search, the
reuse of previous search trees, the time management, and the evaluation
function. \cref{sec:evaluation} presents the experimental evaluation, and
\cref{sec:discussion} concludes with a discussion of the results and directions
for future work.